---
title: "My Analysis"
output: html_document
---

Author: Yos Ramirez

<center>**Thomas Fire**</center>

## Perimeter, True color image, False color image, Map & Air Quality

This code is designed to provide visual context for analyzing wildfire impacts using satellite imagery, combined with air quality data to assess the environmental and health consequences. Specifically this code will look at the Thomas Fire taht occurred in 2017.

### The following code will use data from the sources below

The first dataset contains information a [simplified collection of bands (red, green, blue, near-infrared and shortwave infrared)](https://planetarycomputer.microsoft.com/dataset/landsat-c2-l2).

Microsoft. (n.d.). Landsat Collection 2 Level-2 (C2 L2) data \[Data set\]. Microsoft Planetary Computer. Retrieved November 23, 2024, from https://planetarycomputer.microsoft.com/dataset/landsat-c2-l2

The second dataset is [historical open-access data about fire perimeters in California](https://catalog.data.gov/dataset/california-fire-perimeters-all-b3436) from Data.gov.

U.S. Government. (n.d.). California fire perimeters (ALL) \[Data set\]. Data.gov. Retrieved November 23, 2024, from https://catalog.data.gov/dataset/california-fire-perimeters-all-b3436

[Air Quality Index (AQI)](https://www.airnow.gov/aqi/aqi-basics/) data from the [US Environmental Protection Agency](https://www.epa.gov) to visualize the impact on the AQI of the 2017 [Thomas Fire](https://en.wikipedia.org/wiki/Thomas_Fire) in Santa Barbara County.

## Before we begin, a little on the Thomas fire perimeter

To analyze the Thomas we first need to get the parameters in a shapefile. In a previous .ipynb code chunk I manually selected for the parameter of the Thomas Fire. To briefly explain that code that is not in this blog:

-   I imported essential libraries like requests, zipfile, geopandas, and shutil to download, extract, and manipulate geospatial data

-   I downloaded a ZIP file containing fire perimeter shapefiles from the public URL provided and extract the contents into a temporary directory

-   I loaded the shapefile into a GeoDataFrame and checked the CRS, which I identified as a projected CRS in meters

-   I listed the unique fire names and focused on retrieving the Thomas Fire (2017) by filtering based on the FIRE_NAME and YEAR\_ columns

-   I filtered the dataset to select for the Thomas Fire boundary in 2017, specifically where FIRE_NAME is "THOMAS" and YEAR\_ is 2017

-   I saved the selected perimeter data for the Thomas Fire as a shapefile in a data directory folder, and finally I deleted the temporary extraction directory to keep the workspace clean.

-   The Thomas Fire shapefile is the file I use for this code and is saved as 'Thomas_fire_perimeter.shp'.

## Satellite Imagery that we will explore

-   True color image (RGB)

-   False color image

## True color image (RGB)

What is a True Color Image (Red Green Blue)?

A True Color Image is as it sounds, a visual representation of satellite data where the colors correspond to the red, green, and blue components of the light spectrum, the primary colors of light that the human eye can perceive. When combined, they form a full spectrum of visible colors, with the same idea in mind this is applied to remote sensing and geospatial analysis. Combining data from three specific bands of a satellite image or raster data, creates a true color image. The three bands as you can guess are red, green, and blue, and are very important to plotting the satellite data as you will see in the next few code chunks.

### Bands summary

\- Red Band (corresponds to the red light wavelength, similar to the human eye)

\- Green Band (corresponds to the green light wavelength, similar to the human eye))

\- Blue Band (corresponds to the blue light wavelength, similar to the human eye))

#### Import libraries

```{r, echo=FALSE}
#install.packages("reticulate")
library(reticulate)
# Specify the Python executable path
use_python("/Users/yos/opt/anaconda3/bin/python3")
# Check if Python is configured correctly
py_config()
```

```{r, echo=FALSE}
library(reticulate)
py_run_string("import matplotlib")

```

```{python, echo=FALSE, warning=FALSE}
#| tags: []
import os
import rioxarray as rioxr
import matplotlib.pyplot as plt
import numpy as np
import geopandas as gpd
import fiona
import rasterio
from rasterio.plot import show
```

#### Load in data

```{python}
#| tags: []
# Construct the file path
base_dir = 'data'
landsat = 'landsat8-2018-01-26-sb-simplified.nc'
file_path = os.path.join(base_dir, landsat)
# Open the raster data
landsat_data = rioxr.open_rasterio(file_path)
# Print landsat data
print(landsat_data)
# Print CRS
print("Current CRS:", landsat_data.rio.crs)
landsat_data = landsat_data.rio.reproject("EPSG:3857")
# Update CRS
print("Updated CRS:", landsat_data.rio.crs)
```

#### Explore some key information on the data

```{python}
#| tags: []
# Print coordinates
print("\nDataset Coordinates:",landsat_data.coords)
# Print dimensions
print("\nDataset Dimensions:",landsat_data.dims)
# Print band names and values
print("\nBand Names and Values:",landsat_data.data_vars)
```

After exploration of the data, that is important in determining the next steps to take for the final output of a RGB image, I was able to obtain the following. The data is geospatial data with x, y, which represent the spatial grid and band. The band variables are red, green, blue which we need for a true color image alomng with nir08, and swir22 which we will need in the next part. Each variable has 5mb of data and are float64 or a number that can be a decimal. The dimensions are 889 for the x longitude, and 757 for the y latitude. The crs is WGS 84 or pseudo mercator. The scale factor is 1.0 not requiring alteration.

```{python}
#| tags: []
# Drop single band with squeeze 
landsat_data = landsat_data.squeeze()
# Drop variables associated with band
landsat_data = landsat_data.drop_vars('band')
# Verify dimensions and coordinates
print(landsat_data.dims, landsat_data.coords)
```

-   The squeeze() method is used to remove dimensions of size 1 from an array. This is important when the dataset has dimensions that are not necessary, which could result in confusion or the need for extra processing.

-   The drop_vars() method is used to drop specific variables from a dataset. For this geospatial data it is removing metadata variables, associated with the band previously dropped, that are no longer required.

## Let's look at RGB output

Ensuring the use of `robust=True`

This operation fixes the geometries that may be causing a failure that skew the RGB data, resulting in a fixed true-color image.

The issue here is the clouds, their RGB values are outliers and cause the other values to be squished when plotting.

```{python}
#| tags: []
# Plot landsat data as rbg array, include robust true to fix outliers
landsat_data[['red', 'green', 'blue']].to_array().plot.imshow(robust=True)
plt.title('RGB Image (True Color)') 
plt.show()
```

### Importance of True Color Images for Analyzing Data:

True color images provide a realistic view of the Earth’s surface, a natural representation, similar to what we would see with our own eyes. This can be useful for quickly understanding the landscape, vegetation, water bodies, urban areas, and other natural features. Often true color image is necessary in determining the next spatial analysis to make. As for this analysis of the Thomas Fire, the areas affected by the fire need to be better visualized and so another type of satellite imagery is explored next.

### Summary/Limitations of True Color Images:

-   **Easy visual interpretation** - land cover, as it looks similar to the naked eye.

-   **Cannot See Non-visible Wavelengths** - like infrared, which can be useful for vegetation health analysis.

-   **Cloud Cover** - special processing is needed, such are `robust=True`.

## False color image

Why is it False?

A false color image is a type of image in which other electromagnetic spectrum colors are used to represent different features or properties of the landscape. The colors do not correspond to the colors normally seen with the naked eye. Similar to true color image, false color image needs infrared, short wave infrared, and near infrared bands to highlight specific features.

### Bands Summary

-   Infrared (IR),
-   short-wave infrared (SWIR)
-   near-infrared (NIR)

### Why false color imaging is important

-   **Vegetation Analysis - as we will explore with the Thomas Fire**

#### Example of a false color image

```{python}
#| tags: []
# Plot landsat data as false color image as array, include robust true to fix outliers
landsat_data[['swir22', 'nir08', 'red']].to_array().plot.imshow(robust=True)
plt.title("False Color Image") # Title of plot
plt.show() # Show plot
```

## Time to Map the Thomas Fire

#### Read in data for the perimeter

```{python}
#| tags: []
# Import thomas shape file
fire_shapefile = 'data/Thomas_fire_perimeter.shp' # File path
thomas_fire = gpd.read_file(fire_shapefile) # Read in file with file path
```

#### Make sure CRS is the same

```{python}
# Make crs the same for both data
thomas_fire = thomas_fire.to_crs(landsat_data.rio.crs)
```

### Thomas Fire outlined in red with False Color Image

```{python}
#| tags: []
# Plot Thomas fire parimeter with false color imagery
fig, ax = plt.subplots(figsize=(10, 10)) # Figure size
landsat_data[['swir22','nir08','red']].to_array().plot.imshow(robust=True, # Plot false color from landsat data as array
                                                              ax=ax) # Set axis
thomas_fire.boundary.plot(ax=ax, # Set axis
                          edgecolor='red', # Boundary color
                          linewidth=2, # Boundary width
                          label="Thomas Fire Perimeter") # Boundary label
plt.title("Thomas Fire with False Color Imagery") # Title of plot
plt.legend() # Legend
plt.show() # Show plot
```

#### Figure description

The figure above shows the area of Santa Barbara county which the Thomas fire occurred in 2017, with a red outline for clarity. The image is in short-wave infrared, near-infrared, and red considered to better visualize where the fire occurred. This technique allows for better identification of burned areas. The false color image clearly shows the red land compared to the green, as the NIR band highlights vegetation and areas affected by fire, while the SWIR band distinguishes between different land cover types. Darker red means the area has been burned recently, the lighter red shown above means the area has been burned but not so recent, in 2017.

## Next we will look at the Air Quality Index during the time of the Thomas Fire.

Air quality during and after fires is critically important for several reasons, as wildfires and even smaller fires can release a variety of harmful pollutants into the air, which pose serious health risks to humans and animals. During a wildfire, a number of harmful substances are released into the air. These pollutants can have immediate and long-term health impacts, particularly for vulnerable populations such as children, the elderly, and people with pre-existing health conditions. Wildfires can have a devastating impact on local ecosystems, and poor air quality during and after fires can harm vegetation and animal life.

```{python}
import pandas as pd

aqi_17 = pd.read_csv("data/daily_aqi_by_county_2017.csv")
aqi_18 = pd.read_csv("data/daily_aqi_by_county_2018.csv")

# Combine datasets (2017 and 2018)
aqi = pd.concat([aqi_17, aqi_18])
# Simplify column names
aqi.columns = (aqi.columns
                  .str.lower()
                  .str.replace(' ', '_')
                )
# Save filtered (Santa Barbara) data 
aqi_sb = aqi[aqi['county_name'] == 'Santa Barbara']
# Keep only specific columns (date, aqi, category, defining_parameter, defining_site, and number_of_sites_reporting)
aqi_sb = aqi_sb[['date', 'aqi', 'category', 'defining_parameter', 'defining_site', 'number_of_sites_reporting']]
# Convert specific object column (date) to datetime object
aqi_sb['date'] = pd.to_datetime(aqi_sb['date'])
# Save with new (date) index
aqi_sb = aqi_sb.set_index('date').sort_index()
# Add new 5 day mean column to data (Santa Barbara)
aqi_sb['five_day_average'] = aqi_sb['aqi'].rolling(window='5D').mean()
```

## Plot the AQI highlighting the Thomas Fire dates

```{python}
# Plot figure
plt.figure(figsize=(14, 7))
# Plot aqi , label, color, size, and transparency
plt.plot(aqi_sb.index.values, aqi_sb['aqi'].values, label='Daily AQI', color='darkblue', linewidth=1, alpha=0.7)
# Plot average, label, color, and size
plt.plot(aqi_sb.index.values, aqi_sb['five_day_average'].values, label='5-Day Average AQI', color='fuchsia', linewidth=2)
# Title and size
plt.title('Daily AQI and 5-Day Average AQI (Santa Barbara County)', fontsize=16)
# X label and size
plt.xlabel('Date', fontsize=14)
# Y label and size
plt.ylabel('AQI', fontsize=14)
# Rotate x-axis labels
plt.xticks(rotation=45)
# Legend
plt.legend()
# Grid
plt.grid() 
# Add shaded area for Thomas fire
plt.axvspan(pd.Timestamp('2017-12-01'), pd.Timestamp('2017-12-31'), color='yellow', alpha=0.3)
# Layout
plt.tight_layout()

# Show plot
plt.show()
```

The air quality during the Thomas Fire was high. The Thomas fire shows a spike in the aqi daily shown in blue and 5 day average shown in pink. I've highlighted the month of December yellow to clarify the spike occurence. I suggest doing this to clearly identify the time of interest. It is clear that the daily aqi was much higher than the 5 day average but the 5 day average does remain high for a while compared to normal days. The daily aqi dropped from the highest point to where the average remained. For atleast half of the month of December the daily aqi and the 5 day average were above 150 in which pollution would be a problem. This data visualization shows the importance of being able to gather data and applying the skills of analysis.

Data analysis helps identify the sources and patterns of pollution, whether it's from industrial activities, vehicle emissions, agricultural runoff, or natural events like wildfires. Understanding where and when pollutants are released is the first step in controlling them. By analyzing historical data, we can detect trends in pollution levels, such as increases or decreases in air, water, or soil quality over time. This can help policymakers and environmental agencies make informed decisions about interventions. Data analysis allows for the creation of clear, visual representations of pollution levels, such as maps, graphs, and interactive dashboards. Data analysis can identify areas where pollution control technologies can be most effective, encouraging innovation in clean energy, waste management, or pollution-reducing technologies.

Some information on a few useful libraries

1.  When working with datasets or results that are stored in different directories, `os` helps you manage file paths, allowing your code to be flexible and system-agnostic. For example, os.path.join() is used to create platform-independent file paths.

2.  As for `rioxarray`, this is an extension of the xarray library for handling geospatial raster data (like satellite imagery or other geospatial data formats). It leverages rasterio to read, write, and process raster files.

3.  For visualizations `matplotlib.pyplot` is one of the most commonly used libraries in Python for creating static, interactive, and animated plots and graphs as seen in this blog. In geospatial analysis, you can use plt.imshow() to visualize raster images (such as satellite imagery) and adjust color maps to highlight different features (e.g., vegetation or urban areas).

4.  Geospatial data, rasters and vectors, often require heavy manipulation and `numpy` enables efficient handling of large datasets. As well as performing manipulations on multi-dimensional arrays, which can be common in remote sensing.

<center>**That's it for the Thomas Fire Blog**</center>
